---
title: "Study 1 Data Analysis"
author: "Nicholas Michalak (nickmm@umich.edu)"
date: "06/28/2019"
output: 
  html_document: 
    fig_height: 7.5
    fig_width: 10.5
    theme: flatly
    toc: yes
    toc_float: yes
---

# Install and/or load packages

```{r, warning = FALSE, message = FALSE}

# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("psych")
# install.packages("car")
# install.packages("heplots")

library(tidyverse)
library(knitr)
library(psych)
library(car)
library(heplots)

# I want recode from dplyr
recode <- dplyr::recode

```

# Read data

```{r, warning = FALSE, message = FALSE}

hsb2 <- "data/study1/cleandata/hsb2.csv" %>% read_csv()

```

# Custom function(s)

```{r}

source("functions/oneway.R")

```

# Set factor levels
> I don't typically recommend factors, but sometimes factors are useful.

```{r}

hsb2$ses_program <- factor(hsb2$ses_program, levels = c("Low x General", "Low x Academic", "Low x Vocation", "Middle x General", "Middle x Academic", "Middle x Vocation", "High x General", "High x Academic", "High x Vocation"))

```

# Print contrast variables

```{r}

hsb2 %>% 
  count(ses_program, ses_lbl, ses_linear, ses_quadratic, program_lbl, acadVgenrl, vocVacgen) %>% 
  kable()

```

# Plots

## Does social studies score depend on socioeconomic status and program?

```{r}

hsb2 %>% 
  ggplot(mapping = aes(x = ses_lbl, y = SOCST)) +
  geom_violin(mapping = aes(fill = program_lbl, color = program_lbl), alpha = 0.20) +
  geom_boxplot(mapping = aes(color = program_lbl), position = position_dodge(width = 0.90), width = 0.25, fill = "transparent", outlier.color = NULL) +
  geom_point(mapping = aes(color = program_lbl), position = position_jitterdodge(jitter.width = 0.10, dodge.width = 0.90, seed = 10359), alpha = 0.50) +
  theme_bw()

```

# Model

## Does social studies score depend on socioeconomic status and program?

### Fit linear regression

```{r}

lm1 <- lm(SOCST ~ (ses_linear + ses_quadratic) * (acadVgenrl + vocVacgen), data = hsb2)

```

### Results (specific contrasts)

```{r}

summary(lm1)

```

### ANOVA results

```{r}

Anova(lm1, type = "III")

```

### Eta-Squared and Partial Eta-Squared

```{r}

etasq(lm1, partial = FALSE)
etasq(lm1, partial = TRUE)

```

## ANOVA contrasts, including Welch
> see [Testing Between-Subjects Contrasts in R](https://nickmichalak.com/post/2019-02-13-testing-between-subjects-contrasts-in-r/testing-between-subjects-contrasts-in-r/)

```{r}

# Print levels of grouping variable, ses_program
hsb2 %>% 
  count(ses_program) %>% 
  kable()


with(hsb2,
     oneway(dv = SOCST,
            group = ses_program,
            contrast = list(ses_linear = c(-1, -1, -1, 0, 0, 0, 1, 1, 1) / 3,
                            ses_quad = c(1, 1, 1, -2, -2, -2, 1, 1, 1) / 9,
                            acadVgenrl = c(-1, 1, 0, -1, 1, 0, -1, 1, 0) / 2,
                            vocVacgen = c(-1, -1, 2, -1, -1, 2, -1, -1, 2) / 9),
            alpha = 0.05)
     ) %>% 
  kable()

```

## Does reading aptidude supress, mediate, or confound the relationship between math and science aptitude?
> MacKinnon, D. P., Krull, J. L., & Lockwood, C. M. (2000). Equivalence of the mediation, confounding and suppression effect. Prevention science, 1(4), 173-181. [[.pdf](http://w3dev.ualg.pt/~lfaisca/SMAD05/Prevention%20Science_Mackinnon_2000.pdf)]

### Save math only and math + science models

```{r}

# a path
lm2 <- lm(READ ~ MATH, data = hsb2)

# total effect (c path)
lm3 <- lm(SCIENCE ~ MATH, data = hsb2)

# b path and direct effect (c` path)
lm4 <- lm(SCIENCE ~ MATH + READ, data = hsb2)

```

#### Examine diagnostic plots for the reduced and full model

```{r}

plot(lm3)
plot(lm4)

```

### Results for three models
> 1. READ ~ MATH (a)  
> 2. SCIENCE ~ MATH (total/c)  
> 3. SCIENCE ~ MATH (c`) + READ (b)  
> 4. c - c` = a * b (indirect effect)

```{r}

# a path
summary(lm2)

# total effect (c path)
summary(lm3)

# b path and direct effect (c` path)
summary(lm4)

```

### (Adjusted) R-squared change
> This code is not directly relevant for mediation hypotheses, but it's useful to know for model comparisons.

```{r}

summary(lm4)$adj.r.squared - summary(lm3)$adj.r.squared

```

### Test (adjusted) R-squared change
> You can compare as many models as you want with anova(model1, model2, model3, ...)

```{r}

anova(lm3, lm4)

```

### Simple Indirect Effect Test
> In the analysis script from Day 4, we tested mediation using an SEM framework. This function from the psych package uses a simpler linear regression framework, like the popular Hayes macros.

```{r}

mediate(SCIENCE ~ MATH + (READ), data = hsb2, n.iter = 1000, use = "pairwise", plot = TRUE) %>% 
  print(short = FALSE)

```





