---
title: "study 1 data describe"
date: "06/06/2018"
---

# packages

```{r, warning = FALSE, message = FALSE}

# install.packages("tidyverse")
# install.packages("psych")
# install.packages("knitr")
library(tidyverse)
library(psych)
library(knitr)

```

# data

```{r, warning = FALSE, message = FALSE}

hsb2 <- "data/study1/cleandata/hsb2.csv" %>% read_csv()

```

# demographics

## gender

```{r}

hsb2 %>% 
  count(FEMALE, gender_lbl) %>% 
  kable()

```

## race

```{r}

hsb2 %>% 
  count(RACE, race_lbl) %>% 
  kable()

```

## socioeconomic status

```{r}

hsb2 %>% 
  count(SES, ses_lbl) %>% 
  kable()

```

## school type

```{r}

hsb2 %>% 
  count(SCHTYP, schooltype_lbl) %>% 
  kable()

```

## program

```{r}

hsb2 %>% 
  count(PROG, program_lbl) %>% 
  kable()

```

# visualize reading, writing, math, and science scores

## boxplots of scores by gender

```{r}

hsb2 %>% 
  select(gender_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(x = gender_lbl, y = score)) +
  geom_boxplot() +
  facet_wrap(facets = ~ measure)

```

## boxplots of scores by race

```{r}

hsb2 %>% 
  select(race_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(x = race_lbl, y = score)) +
  geom_boxplot() +
  facet_wrap(facets = ~ measure)

```

## boxplots of scores by socioeconomic status

```{r}

hsb2 %>% 
  select(ses_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(x = ses_lbl, y = score)) +
  geom_boxplot() +
  facet_wrap(facets = ~ measure)

```

## boxplots of scores by school type

```{r}

hsb2 %>% 
  select(schooltype_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(x = schooltype_lbl, y = score)) +
  geom_boxplot() +
  facet_wrap(facets = ~ measure)

```

## boxplots of scores by program

```{r}

hsb2 %>% 
  select(program_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(x = program_lbl, y = score)) +
  geom_boxplot() +
  facet_wrap(facets = ~ measure)

```

# visualize reading, writing, math, and science score normality

## QQ-plots by gender

```{r}

hsb2 %>% 
  select(gender_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(sample = score)) +
  geom_qq() +
  facet_grid(facets = gender_lbl ~ measure)

```

## QQ-plots by race

```{r}

hsb2 %>% 
  select(race_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(sample = score)) +
  geom_qq() +
  facet_grid(facets = race_lbl ~ measure)

```

## QQ-plots by socioeconomic status

```{r}

hsb2 %>% 
  select(ses_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(sample = score)) +
  geom_qq() +
  facet_grid(facets = ses_lbl ~ measure)

```

## QQ-plots by school type

```{r}

hsb2 %>% 
  select(schooltype_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(sample = score)) +
  geom_qq() +
  facet_grid(facets = schooltype_lbl ~ measure)

```

## QQ-plots by program

```{r}

hsb2 %>% 
  select(program_lbl, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  gather(key = measure, value = score, READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  ggplot(mapping = aes(sample = score)) +
  geom_qq() +
  facet_grid(facets = program_lbl ~ measure)

```

# scatterplot matrices (i.e., correlations and histograms)

## ignoring gender, race, socioeconomic status, schooltype, and program

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  pairs.panels(scale = FALSE, smooth = TRUE, pch = ".", ci = TRUE)

```

# describe reading, writing, math, and science scores

## describe by gender

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  describeBy(group = select(hsb2, gender_lbl), mat = TRUE) %>% 
  kable()

```

## describe by race

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  describeBy(group = select(hsb2, race_lbl), mat = TRUE) %>% 
  kable()

```

## describe by socioeconomic status

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  describeBy(group = select(hsb2, ses_lbl), mat = TRUE) %>% 
  kable()

```

## describe by school type

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  describeBy(group = select(hsb2, schooltype_lbl), mat = TRUE) %>% 
  kable()

```

## describe by program

```{r}

hsb2 %>% 
  select(READ, WRITE, MATH, SCIENCE, SOCST) %>% 
  describeBy(group = select(hsb2, program_lbl), mat = TRUE) %>% 
  kable()

```
