---
title: "Day 3. Vectors, strings, factors, and statistics"
author: "Nicholas Michalak"
date: "6/24/2017"
output: 
  html_document:
    fig_height: 7.5
    fig_width: 10.5
    keep_md: yes
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

# install prerequisite packages

```{r}

# character vector of packages you'll need for your whole analysis
needed_packages <- c("tidyverse", "forcats", "psych", "haven", "lavaan", "lme4", "lmerTest", "afex", "compute.es")

# source custom function
lapply(list.files(path = "custom_functions/", pattern = "*.R", full.names = TRUE), source)

# install needed packages
install_needed_pkgs(needed_packages = needed_packages)

```

# integer
> subject ID

```{r}

(subj_id <- 1:20)

```

# logical
> did this subject fail the pre-screen?

```{r}

(failed_prescreen <- c(TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE))

```

# character / factor
> experimental group

```{r}

# character vector
(exp_group <- c("Treatment", "Control", "Control", "Treatment", "Treatment", "Control", "Treatment", "Control", "Treatment", "Treatment", "Treatment", "Control", "Control", "Treatment", "Control", "Control", "Treatment", "Control", "Treatment", "Control"))

# convert to factor vector
(exp_group_fac <- parse_factor(exp_group,
                               levels = c("Treatment", "Control")))

```

## modify factor order

```{r}

(exp_group_fac <- parse_factor(exp_group,
                               levels = c("Control", "Treatment")))

```

## modify factor levels

```{r}

(exp_group_fac <- recode_factor(exp_group_fac, Control = "Neutral", Treatment = "Manipulation"))

```

# numeric
> dependent variable

```{r}

(dep_var <- c(4, 2, 7, 3, 5, 6, 6, 1, 7, 3, 4, 1, 3, 6, 1, 7, 4, 7, 5, 1))

```

# combine them into a "tibble"

```{r}

(my_data <- tibble(subj_id, exp_group, exp_group_fac, dep_var, failed_prescreen))

```

# review

## `summarise()`

```{r}

my_data %>%
  group_by(failed_prescreen) %>%
  summarise(n = n())

```

## `count()`

```{r}

my_data %>%
  count(failed_prescreen)

```

## `filter()`

```{r}

my_data %>%
  filter(failed_prescreen == FALSE)

```

## `geom_histogram()` and aesetics

```{r}

my_data %>%
  ggplot(mapping = aes(x = dep_var, fill = failed_prescreen)) +
  geom_histogram(binwidth = 1, color = "white")

```

### `facet_wrap()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = dep_var)) +
  geom_histogram(binwidth = 1, color = "white") +
  facet_wrap(~ failed_prescreen)

```

## `geom_boxplot()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  geom_boxplot()

```

## `geom_violin()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  geom_violin()

```

## `stat_summary()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "bar", position = position_dodge(0.9)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.1, position = position_dodge(0.9))

```

## grouped descriptives

```{r}

with(data = my_data, describeBy(dep_var, group = exp_group_fac, mat = TRUE))

```

## `t.test()`
* I make all arguments explcit

```{r, error = TRUE}

t.test(formula = dep_var ~ exp_group_fac,
       data = my_data,
       alternative = "two.sided",
       mu = 0,
       paired = FALSE,
       var.equal = FALSE,
       conf.level = .95)

```

### less explicit
* if you don't name arguments explictly, make sure you give argument input in order

```{r}

t.test(dep_var ~ exp_group_fac, my_data)

```

# Big Five data from [personality-testing.info](http://personality-testing.info/_rawdata/)

```{r}

(big_five <- "example_tidy_data/big_five_may_2014/big_five_may_2014.csv" %>%
  read_tsv() %>%
  mutate(race_fac = recode(race, `1`  =  "Mixed Race", `2`  =  "Arctic (Siberian, Eskimo)", `3`  =  "Caucasian (European)", `4` = "Caucasian (Indian)", `5` = "Caucasian (Middle East)", `6` = "Caucasian (North African, Other)", `7` = "Indigenous Australian", `8` = "Native American", `9` = "North East Asian (Mongol, Tibetan, Korean Japanese, etc)", `10` = "Pacific (Polynesian, Micronesian, etc)", `11` = "South East Asian (Chinese, Thai, Malay, Filipino, etc)", `12` = "West African, Bushmen, Ethiopian", `13` = "Other (0 = missed)"),
         eng_nat_fac = recode(engnat, `1` = "yes", `2` = "no", `0` = "missed"),
         gender_fac = recode(gender, `1` = "Male", `2` = "Female", `3` = "Other", `0` = "missed"),
         hand_fac = recode(hand, `1` = "Right", `2` = "Left", `3` = "Both", `0` = "missed"),
         source_fac = recode(source, `1` = "from another page on the test website", `2` = "from google", `3` = "from facebook", `4` = 'from any url with ".edu" in its domain name (e.g. xxx.edu, xxx.edu.au)', `5` = "other source, or HTTP Referer not provided")) %>%
  select(age, country, race, race_fac, engnat, eng_nat_fac, gender, gender_fac, hand, hand_fac, source, source_fac, everything(.)))

```

# more descriptives examples

* see psych package ([an overview of the psych package](http://personality-project.org/r/overview.pdf)) for many trusty tools

## extraversion

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  psych::describe(.)
  
```

# some correlation matrix examples

## `pairs.panels()`

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  pairs.panels(pch = ".")

```

## `cor.plot()`

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  cor.plot(numbers = TRUE)

```

## `cor.test()`
* explicit argument names again

```{r}

big_five %>%
  cor.test(formula = ~ E1 + E2, data = .,
           alternative = "two.sided",
           method = "pearson",
           exact = NULL,
           conf.level = .95)

```

## `corr.test()`
* from psych package
* takes matrix or dataframe input
* explicit argument names again

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  corr.test(use = "complete",
            method = "pearson",
            adjust = "none",
            alpha = .05,
            ci = TRUE) %>%
  print(short = FALSE)

```

# regression

## `lm()`
* to be honest, I don't understand *all* of the arguments in `lm()`

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3, data = .)
  

```

## `summary()`

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3, data = .) %>%
  summary(.)

```

### interactions
* ":" creates an interaction term between variables
* "*" creates all main effect and interaction terms

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3 + E2:E3, data = .) %>%
  summary(.)

# same model
big_five %>%
  lm(formula = E1 ~ E2 * E3, data = .) %>%
  summary(.)

```

## `glm()`

```{r}

big_five %>%
  filter(hand != 3 & hand != 0) %>%
  mutate(right_handed = recode(hand, `1` = 1, `2` = 0)) %>%
  glm(formula = right_handed ~ E2 * E3, data = ., family = binomial) %>%
  summary(.)

```

# Factorial ANOVA

## afex package

### between-subjects ANOVA

```{r}

my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  summary(.)

```

### specific contrasts

```{r}

# t-value and p-value
my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  lsmeans::lsmeans(specs = "exp_group_fac") %>%
  contrast(list(ntrl_mnpl = c(-1, 1)))

# confidence intervals
my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  lsmeans::lsmeans(specs = "exp_group_fac") %>%
  contrast(list(ntrl_mnpl = c(-1, 1))) %>%
  confint(.)

```

## straight to contrasts

### Function 23: Confidence interval for a linear contrast of means (between-subjects design) [[Download more functions from Douglas Bonnet's course webpage](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjg7bG3u-bUAhVi2IMKHWfSAWwQFggqMAA&url=https%3A%2F%2Fpeople.ucsc.edu%2F~dgbonett%2Fdocs%2Fpsyc204%2F204RFunctions.docx&usg=AFQjCNGRCAg-lE7TwzMSps_A1pmDOEEb8w)]

```{r}

my_data %>%
  group_by(exp_group_fac) %>%
  summarise(n = n(),
            m = mean(dep_var),
            sd = sd(dep_var)) %>%
  with(data = ., CImeanBS(alpha = .05, m = m, sd = sd, n = n, c = c(-1, 1)))

```

## effect size

### Function 24: Confidence interval for a standardized linear contrast of means (between-subjects design) [[Download more functions from Douglas Bonnet's course webpage](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjg7bG3u-bUAhVi2IMKHWfSAWwQFggqMAA&url=https%3A%2F%2Fpeople.ucsc.edu%2F~dgbonett%2Fdocs%2Fpsyc204%2F204RFunctions.docx&usg=AFQjCNGRCAg-lE7TwzMSps_A1pmDOEEb8w)]

```{r}

my_data %>%
  group_by(exp_group_fac) %>%
  summarise(n = n(),
            m = mean(dep_var),
            sd = sd(dep_var)) %>%
  with(data = ., CIstdmeanBS(alpha = .05, m = m, sd = sd, n = n, c = c(-1, 1)))


```

