---
title: "Day 3. Vectors, strings, factors, and statistics"
author: "Nicholas Michalak"
date: "6/24/2017"
output: 
  html_document:
    fig_height: 7.5
    fig_width: 10.5
    keep_md: yes
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

# install prerequisite packages

```{r}

# character vector of packages you'll need for your whole analysis
needed_packages <- c("tidyverse", "magrittr", "forcats", "psych", "haven", "lavaan", "lme4", "lmerTest", "afex", "compute.es", "broom")

# source custom function
lapply(list.files(path = "custom_functions/", pattern = "*.R", full.names = TRUE), source)

# install needed packages
install_needed_pkgs(needed_packages = needed_packages)

```

# integer
> subject ID

```{r}

(subj_id <- 1:20)

```

# logical
> did this subject fail the pre-screen?

```{r}

(failed_prescreen <- c(TRUE, FALSE, FALSE, FALSE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE, FALSE))

```

# character / factor
> experimental group

```{r}

# character vector
(exp_group <- c("Treatment", "Control", "Control", "Treatment", "Treatment", "Control", "Treatment", "Control", "Treatment", "Treatment", "Treatment", "Control", "Control", "Treatment", "Control", "Control", "Treatment", "Control", "Treatment", "Control"))

# convert to factor vector
(exp_group_fac <- parse_factor(exp_group,
                               levels = c("Treatment", "Control")))

```

## modify factor order

```{r}

(exp_group_fac <- parse_factor(exp_group,
                               levels = c("Control", "Treatment")))

```

## modify factor levels

```{r}

(exp_group_fac <- recode_factor(exp_group_fac, Control = "Neutral", Treatment = "Manipulation"))

```

# numeric
> dependent variable

```{r}

(dep_var <- c(4, 2, 7, 3, 5, 6, 6, 1, 7, 3, 4, 1, 3, 6, 1, 7, 4, 7, 5, 1))

```

# combine them into a "tibble"

```{r}

(my_data <- tibble(subj_id, exp_group, exp_group_fac, dep_var, failed_prescreen))

```

# review

## `summarise()`

```{r}

my_data %>%
  group_by(failed_prescreen) %>%
  summarise(n = n())

```

## `count()`

```{r}

my_data %>%
  count(failed_prescreen)

```

## `filter()`

```{r}

my_data %>%
  filter(failed_prescreen == FALSE)

```

## `geom_histogram()` and aesetics

```{r}

my_data %>%
  ggplot(mapping = aes(x = dep_var, fill = failed_prescreen)) +
  geom_histogram(binwidth = 1, color = "white")

```

### `facet_wrap()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = dep_var)) +
  geom_histogram(binwidth = 1, color = "white") +
  facet_wrap(~ failed_prescreen)

```

## `geom_boxplot()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  geom_boxplot()

```

## `geom_violin()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  geom_violin()

```

## `stat_summary()`

```{r}

my_data %>%
  ggplot(mapping = aes(x = exp_group_fac, y = dep_var, fill = exp_group_fac)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "bar", position = position_dodge(0.9)) +
  stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.1, position = position_dodge(0.9))

```

## grouped descriptives
> Note: `%$%` is called the "exposition pipe operator." From the [magrittr documentation](https://cran.r-project.org/web/packages/magrittr/magrittr.pdf) description, "Some functions, e.g. lm and aggregate, have a data argument, which allows the direct use of names inside the data as part of the call. This operator exposes the contents of the left-hand side object to the expression on the right to give a similar benefit, see the examples."

### quick `%$%` example (where `%>%` wouldn't work)

```{r}

iris %$%
  cor(Sepal.Length, Sepal.Width)

```

### now to grouped descriptives

```{r}

my_data %$%
  describeBy(x = dep_var, group = exp_group_fac, mat = TRUE)

```

### cite your tools

```{r}

citation("psych")

```

## `t.test()`
* I make all arguments explcit

```{r, error = TRUE}

t.test(formula = dep_var ~ exp_group_fac,
       data = my_data,
       alternative = "two.sided",
       mu = 0,
       paired = FALSE,
       var.equal = FALSE,
       conf.level = .95)

```

### less explicit
* if you don't name arguments explictly, make sure you give argument input in order

```{r}

t.test(dep_var ~ exp_group_fac, my_data)

```

# Big Five data from [personality-testing.info](http://personality-testing.info/_rawdata/)

```{r}

(big_five <- "example_tidy_data/big_five_may_2014/big_five_may_2014.csv" %>%
  read_tsv() %>%
  mutate(race_fac = recode(race, `1`  =  "Mixed Race", `2`  =  "Arctic (Siberian, Eskimo)", `3`  =  "Caucasian (European)", `4` = "Caucasian (Indian)", `5` = "Caucasian (Middle East)", `6` = "Caucasian (North African, Other)", `7` = "Indigenous Australian", `8` = "Native American", `9` = "North East Asian (Mongol, Tibetan, Korean Japanese, etc)", `10` = "Pacific (Polynesian, Micronesian, etc)", `11` = "South East Asian (Chinese, Thai, Malay, Filipino, etc)", `12` = "West African, Bushmen, Ethiopian", `13` = "Other (0 = missed)"),
         eng_nat_fac = recode(engnat, `1` = "yes", `2` = "no", `0` = "missed"),
         gender_fac = recode(gender, `1` = "Male", `2` = "Female", `3` = "Other", `0` = "missed"),
         hand_fac = recode(hand, `1` = "Right", `2` = "Left", `3` = "Both", `0` = "missed"),
         source_fac = recode(source, `1` = "from another page on the test website", `2` = "from google", `3` = "from facebook", `4` = 'from any url with ".edu" in its domain name (e.g. xxx.edu, xxx.edu.au)', `5` = "other source, or HTTP Referer not provided")) %>%
  select(age, country, race, race_fac, engnat, eng_nat_fac, gender, gender_fac, hand, hand_fac, source, source_fac, everything()))

```

# more descriptives examples

* see psych package ([an overview of the psych package](http://personality-project.org/r/overview.pdf)) for many trusty tools

## extraversion

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  psych::describe()
  
```

# some correlation matrix examples

## `pairs.panels()`

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  pairs.panels(pch = ".")

```

## `cor.plot()`

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  cor.plot(numbers = TRUE)

```

## `cor.test()`
* explicit argument names again

```{r}

big_five %$%
  cor.test(E1, E2,
           alternative = "two.sided",
           method = "pearson",
           exact = NULL,
           conf.level = .95)

```

## `corr.test()`
* from psych package
* takes matrix or dataframe input
* explicit argument names again

```{r}

big_five %>%
  select(num_range(prefix = "E", range = 1:10)) %>%
  corr.test(use = "complete",
            method = "pearson",
            adjust = "none",
            alpha = .05,
            ci = TRUE) %>%
  print(short = FALSE)

```

# regression

## `lm()`
* to be honest, I don't understand *all* of the arguments in `lm()`

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3, data = .)

```

## `summary()`

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3, data = .) %>%
  summary()

```

### interactions
* ":" creates an interaction term between variables
* "*" creates all main effect and interaction terms

```{r}

big_five %>%
  lm(formula = E1 ~ E2 + E3 + E2:E3, data = .) %>%
  summary()

# same model
big_five %>%
  lm(formula = E1 ~ E2 * E3, data = .) %>%
  summary()

```

## `glm()`

```{r}

big_five %>%
  filter(hand != 3 & hand != 0) %>%
  mutate(right_handed = recode(hand, `1` = 1, `2` = 0)) %>%
  glm(formula = right_handed ~ E2 * E3, data = ., family = binomial) %>%
  summary()

```

# Factorial ANOVA

## afex package

### between-subjects ANOVA

```{r}

my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  summary()

```

### specific contrasts

```{r}

# t-value and p-value
my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  lsmeans::lsmeans(specs = "exp_group_fac") %>%
  contrast(list(ntrl_mnpl = c(-1, 1)))

# confidence intervals
my_data %>%
  aov_car(formula = dep_var ~ exp_group_fac + Error(subj_id), data = .) %>%
  lsmeans::lsmeans(specs = "exp_group_fac") %>%
  contrast(list(ntrl_mnpl = c(-1, 1))) %>%
  confint()

```

### cite your tools

```{r}

citation("afex")

```

## straight to contrasts

### Function 23: Confidence interval for a linear contrast of means (between-subjects design) [[Download more functions from Douglas Bonnet's course webpage](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjg7bG3u-bUAhVi2IMKHWfSAWwQFggqMAA&url=https%3A%2F%2Fpeople.ucsc.edu%2F~dgbonett%2Fdocs%2Fpsyc204%2F204RFunctions.docx&usg=AFQjCNGRCAg-lE7TwzMSps_A1pmDOEEb8w)]

#### D. Bonnet's example

```{r}

m <- c(33.5, 37.9, 38.0, 44.1)
sd <- c(3.84, 3.84, 3.65, 4.98)
n <- c(10, 10, 10, 10)
c <- c(0.5, 0.5, -0.5, -0.5)
CIstdmeanBS(alpha = .05, m = m, sd = sd, n = n, c = c)

```

#### example using `summarise()` and pipes

```{r}

my_data %>%
  group_by(exp_group_fac) %>%
  summarise(n = n(),
            m = mean(dep_var),
            sd = sd(dep_var)) %$%
  CImeanBS(alpha = .05, m = m, sd = sd, n = n, c = c(-1, 1))

```

## effect size

### Function 24: Confidence interval for a standardized linear contrast of means (between-subjects design) [[Download more functions from Douglas Bonnet's course webpage](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&ved=0ahUKEwjg7bG3u-bUAhVi2IMKHWfSAWwQFggqMAA&url=https%3A%2F%2Fpeople.ucsc.edu%2F~dgbonett%2Fdocs%2Fpsyc204%2F204RFunctions.docx&usg=AFQjCNGRCAg-lE7TwzMSps_A1pmDOEEb8w)]

#### D. Bonnet's example: just enter arguments as vector objects

```{r}

m <- c(33.5, 37.9, 38.0, 44.1)
sd <- c(3.84, 3.84, 3.65, 4.98)
n <- c(10, 10, 10, 10)
c <- c(0.5, 0.5, -0.5, -0.5)
CIstdmeanBS(alpha = .05, m = m, sd = sd, n = n, c = c)

```

#### example using `summarise()` and pipes

```{r}

my_data %>%
  group_by(exp_group_fac) %>%
  summarise(n = n(),
            m = mean(dep_var),
            sd = sd(dep_var)) %$%
  CIstdmeanBS(alpha = .05, m = m, sd = sd, n = n, c = c(-1, 1))


```

# Linear Mixed Effects Models

## lme4 package

### sleepstudy data (comes with lme4)
> The average reaction time per day for subjects in a sleep deprivation study. On day 0 the subjects had their normal amount of sleep. Starting that night they were restricted to 3 hours of sleep per night. The observations represent the average reaction time on a series of tests given each day to each subject.
* Reaction: Average reaction time (ms)
* Days: Number of days of sleep deprivation
* Subject: Subject number on which the observation was made.
> Note: I'm borrowing from a dizzyingly insightful blogpost by Tristan Mahr at UW-Madison [Plotting partial pooling in mixed-effects models](https://tjmahr.github.io/plotting-partial-pooling-in-mixed-effects-models/)

```{r}

?sleepstudy

```

### fit model and summarise

```{r}

sleepstudy %>%
  lmer(Reaction ~ Days + (1 + Days | Subject), data = ., REML = TRUE) %>%
  summary()

```

### scatterplot *Reaction* on *Days* for each *Subject*

```{r}

sleepstudy %>%
  ggplot(aes(x = Days, y = Reaction)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ Subject, nrow = 3)

```

### use `lmList()` to get those intercepts and slopes you see above (no pooling / estimates ignore other participants)

```{r}

(no_pooling <- 
sleepstudy %>%
  lmList(Reaction ~ Days | Subject, data = .) %>% 
  coefficients() %>%
  as_tibble() %>%
  rownames_to_column(var = "Subject") %>%
  rename(Intercept = `(Intercept)`,
         Slope = Days) %>%
  mutate(model = "No pooling"))

```

### use `lm()` to get slope and intercept (full pooling / each participant gets the same value)

```{r}

(complete_pooling <- 
tibble(Subject = sleepstudy$Subject %>% unique() %>% parse_character(),
       Intercept = sleepstudy %>% lm(Reaction ~ Days, data = .) %>% coefficients() %>% .[["(Intercept)"]],
       Slope = sleepstudy %>% lm(Reaction ~ Days, data = .) %>% coefficients() %>% .[["Days"]]) %>%
  mutate(model = "Complete pooling"))

```

### use `lmer()` to get slopes and intercepts for each participant (partial pooling / uses individual estimates adjusted by what's known about other participants)

```{r}

(partial_pooling <- 
sleepstudy %>%
  lmer(Reaction ~ Days + (1 + Days | Subject), data = ., REML = TRUE) %>%
  coefficients() %>%
  .[["Subject"]] %>%
  rownames_to_column(var = "Subject") %>%
  as_tibble() %>%
  rename(Intercept = `(Intercept)`,
         Slope = Days) %>%
  mutate(model = "Partial pooling"))

```

### plot these estimates

```{r}

bind_rows(no_pooling, partial_pooling, complete_pooling) %>% 
  left_join(sleepstudy, by = "Subject") %>%
  ggplot(aes(x = Days, y = Reaction)) +
  geom_point() +
  geom_abline(aes(intercept = Intercept, slope = Slope, color = model)) +
  scale_x_continuous(breaks = seq(0, 9, 1), limits = c(0, 9)) +
  facet_wrap(~ Subject, nrow = 3) +
  theme(legend.position = "top")

```

### cite your tools

```{r}

citation("lme4")
citation("lmerTest")

```

# structural equation models

## lavaan package

### mediation example
> "The data for the 50 subjects in this hypothetical study of the effects of room temperature on water consumption are shown in Table 3.1, where X is the temperature in degrees Fahrenheit, M is a self-report measure of thirst at the end of a 2-hour period, and Y is the number of deciliters of water consumed during the last 2 hours of the study. The 50 subjects were in a room for 4 hours doing a variety of tasks including sorting objects, tracking objects on a computer screen, and communicating via an intercom system. The tasks were selected to represent activities of persons alone in contained environments such as a submarine, spacecraft, or space suit. Before the experiment, each participant was acclimated to a standard temperature of 70°F. Temperature, the independent variable, was then manipulated such that each participant was exposed to a specifi c temperature in the room for the 4 hours of the experiment. At the end of 2 hours, the subjects reported how thirsty they were on a 1 to 5 scale from 1 (not at all thirsty) to 5 (very thirsty). During the last 2 hours of the experiment, water was made available in the room, and the number of the deciliters of water the subjects drank was recorded. (MacKinnon, 2008, p. 55-56)" 

```{r}

subj_num <- 1:50

x <- c(70, 71, 69, 70, 71, 70, 69, 70, 70, 72, 71, 71, 70, 71, 71, 70, 70, 69, 72, 71, 71, 72, 67, 71, 71, 70, 70, 69, 69, 70, 71, 70, 70, 70, 71, 68, 72, 69, 70, 68, 68, 70, 71, 69, 69, 71, 71, 71, 72, 70)

m <- c(4, 4, 1, 1, 3, 4, 3, 5, 4, 5, 2, 3, 5, 4, 4, 2, 4, 3, 3, 3, 2, 3, 1, 4, 3, 3, 2, 3, 4, 3, 2, 1, 2, 2, 4, 2, 4, 3, 3, 3, 3, 4, 4, 2, 3, 3, 4, 3, 4, 2)

y <- c(3, 3, 3, 3, 3, 2, 3, 5, 4, 4, 2, 4, 5, 5, 5, 2, 4, 5, 4, 3, 4, 5, 2, 4, 2, 4, 3, 4, 3, 3, 1, 3, 5, 1, 3, 1, 3, 2, 3, 2, 3, 3, 4, 2, 3, 4, 4, 2, 5, 2)

table_3_1 <- tibble(subj_num, x, m, y)


```

#### write model

```{r}

mediation_model <- "# a path
                      m ~ a * x

                    # b path
                      y ~ b * m

                    # c path (direct effect)
                      y ~ c * x

                    # print total and indirect effects
                      ab := a * b
                      total := c + (a * b)"

```

#### fit model with `sem()`

```{r}

mediation_model %>%
  sem(data = table_3_1)

```

#### summarise model

```{r}

mediation_model %>%
  sem(data = table_3_1) %>%
  summary()

```

#### parameter estimates

```{r}

mediation_model %>%
  sem(data = table_3_1,
      se = "bootstrap",
      bootstrap = 1000) %>%
  parameterestimates(boot.ci.type = "bca.simple")

```

### confirmatory factor analysis example from lavaan website [A CFA example](http://lavaan.ugent.be/tutorial/cfa.html)
> The classic Holzinger and Swineford (1939) dataset consists of mental ability test scores of seventh- and eighth-grade children from two different schools (Pasteur and Grant-White). In the original dataset (available in the MBESS package), there are scores for 26 tests. However, a smaller subset with 9 variables is more widely used in the literature (for example in Joreskog's 1969 paper, which also uses the 145 subjects from the Grant-White school only).

```{r}

?HolzingerSwineford1939

```

#### write model

```{r}

cfa_model <- "# latent variables
                visual  =~ x1 + x2 + x3 
                textual =~ x4 + x5 + x6
                speed   =~ x7 + x8 + x9"

```

#### fit model with `cfa()`

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939)

```

#### summarise model

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  summary()

```

#### fit measures

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  fitmeasures() %>%
  tidy() %>%
  mutate(value = round(x, 4)) %>%
  print(n = 42)

```

#### parameter estimates

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  parameterestimates()

```

#### modification indices

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  modificationindices()

```

#### implied covariance matrix

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  fitted()

```

#### residual matrix

```{r}

cfa_model %>%
  cfa(data = HolzingerSwineford1939) %>%
  residuals()

```

### cite your tools

```{r}

citation("lavaan")

```

