---
title: "Day 2. Workflow and data transformation"
author: "Iris Wang"
date: "June 23, 2017"
output: 
  html_document:
    fig_height: 7.5
    fig_width: 10.5
    keep_md: yes
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 20px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Basics <br>
## 1. Basic functionality (Tools ----> Keyboard shortcuts help) <br>
  + You can use the up arrow key to replay a line you already ran
  + Ctrl Enter will run a line
  + Alt +- will draw an arrow
  + Ctrl Shift M will draw a pipe operator
  + R is case sensitive so it matters if you caps or not, have a space in there or not
  + When writing code, use # in front of a comment. Frequently comment your code so you know what's happening 5 years later.
  
## 2. R is like a calculator
  + Operators are the same (+, -, *, /)
  
```{r}

1 + 1

2 - 4

5 * 8

8 / 4

```

## 3. Create new objects with <-
  + Object name <- value
  
```{r}

# save
x <- 3 * 4

# print
x

# save
y <- 10 / 2

# print
y

# save
z <- x * y

# print
z

```

  + Names should start w/ a character and be helpful (i.e., not a, b, c, but real descriptors of what the thing is)
  
## 4. Calling functions
  + R has several built-in functions
  + Sometimes a function you need will be in a specific package
      + Installing and Loading packages
      
```{r}

# function for only installing packages you don't have and loading libraries you need
source("custom_functions/install_needed_pkgs.R")

# character vector for needed packages
needed_packages <- c("tidyverse")

install_needed_pkgs(needed_packages)

```

  + ?function will bring up a help screen for that function
      + Can see what the arguments are that you need
      + Can see the defaults for the arguments
  + Calling a function requires you type the name + the arguments it requires 
      + Demo: merge(df1, df2, by=”variable name”, all=TRUE)
      
```{r}
# create four variables: N, id, cond x, z, and y
# 100 observations per variable
N <- 100

# create a subject id 
id <- 1:100

# create a condition variable
cond <- rep(c("Treatment", "Control"), each = 50)

# set randomizer seed so results can be reproduced
set.seed(1234)

# drawn from random normal distribution
gratitude_t1 <- rnorm(n = N, mean = 4.5, sd = 0.95)
gratitude_t2 <- rnorm(n = N, mean = 5.5, sd = 0.95)
interdep <- 0.8 * x + rnorm(n = N, mean = 0, sd = 0.5)

# store in a dataframe
time1_survey <- tibble(id, cond, gratitude_t1)
time2_survey <- tibble(id, gratitude_t2, interdep)

# join time 1 and time 2 by subject id
# ?left_join (Join matching rows from y to x)
# right_join
# ?full_join
joined_survey <- left_join(x = time1_survey, y = time2_survey, by = "id")

```
      
      + Demo: scale(variable, scale=FALSE)
        + scale will center your variable if the argument scale is FALSE
        + it will z-score your variable if the argument scale is TRUE
        
```{r}

# ?scale
joined_survey <- joined_survey %>%
  mutate(interdep_c = as.numeric(scale(interdep, center = TRUE, scale = FALSE)))

```

## 5. Useful functions for data management/transformation
  + filter()
  
```{r}

# filter only id "exactly equal to" 90
dplyr::filter(joined_survey, id == 90)

# filter only id "less than" 90
dplyr::filter(joined_survey, id < 90)

# filter only id "less than" 9 "or" id "exactly equal to" 14
dplyr::filter(joined_survey, id < 9 | id == 14)

# filter only id "less than" 9 "and" gratitude_t1 "greater than" 5
dplyr::filter(joined_survey, id < 9 & gratitude_t1 > 5)

# filter only condition "not equal to" Treatment
dplyr::filter(joined_survey, cond != "Treatment")

```
  
  + arrange()
    + arrange allows you to sort the data based on certain columns
    
```{r}

# arrange joined_survey by cond
arrange(joined_survey, cond)

# arrange joined_survey by id
arrange(joined_survey, id)

```
  
  + select()
    + select allows you to select specific columns (for instance if you're working with a giant dataframe)
    
```{r}

# from joined_survey, select id, interdep_c, and gratitude_t1
select(joined_survey, id, interdep_c, gratitude_t1)

```
  
  + mutate()
    + mutate allows you to create new variables
    
```{r}

# mutate gratitude_mean
# two ways to do it
# note: not saving mutated variables; to do so, must save to old or new object (joined_survey) with <-
mutate(joined_survey,
       gratitude_mean = (gratitude_t1 + gratitude_t2) / 2,
       gratitude_mean_alt = mean(c(gratitude_t1, gratitude_t2),
                                 na.rm = TRUE)
)

# using pipes
# have to use rowwise
joined_survey %>% 
  rowwise(.) %>% 
  mutate(gratitude_mean = (gratitude_t1 + gratitude_t2)/2, 
         gratitude_mean_alt= mean(c(gratitude_t1, gratitude_t2),
                                  na.rm = TRUE)
  )
```
  
  + summarise()
    + summarise() will also allow you to create variables at the total group level
    
```{r}

summarise(joined_survey, mean(x = c(gratitude_t1, gratitude_t2), na.rm = TRUE))

```
   
   + gather()
    + puts data into long format
    
```{r}

# variable name (key) = time and response value (value) = gratitude
# arrange by Subj ID
joined_survey_long <- gather(data = joined_survey, key = time, value = gratitude, gratitude_t1, gratitude_t2) %>%
  arrange(id)

```
 
 + spread() 
    + puts data into wide format
    
```{r}

joined_survey_wide <- spread(data = joined_survey_long, key = time, value = gratitude)

```
