---
title: "Day 2. Workflow and data transformation"
author: "Iris Wang"
date: "June 23, 2017"
output: 
  html_document:
    fig_height: 7.5
    fig_width: 10.5
    keep_md: yes
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<style type="text/css">

body{ /* Normal  */
      font-size: 20px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 38px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 28px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 22px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>

# Basics <br>
## 1. Basic functionality (Tools ----> Keyboard shortcuts help) <br>
  + You can use the up arrow key to replay a line you already ran
  + Ctrl Enter will run a line
  + Alt +- will draw an arrow
  + Ctrl Shift M will draw a pipe operator
  + R is case sensitive so it matters if you caps or not, have a space in there or not
  + When writing code, use # in front of a comment. Frequently comment your code so you know what's happening 5 years later.
  + broom thing will always clear environment stuff
  
## 2. R is like a calculator
  + Operators are the same (+, -, *, /)
  
```{r}

1 + 1

2 - 4

5 * 8

8 / 4

```

## 3. Create new objects with `<-`
  + Object name <- value
  
```{r}

# save
x <- 3 * 4

# print
x

# save
y <- 10 / 2

# print
y

# save
z <- x * y

# print
z

#rm() removes objects from your memory
rm(x, y)

```

  + Names should start w/ a character and be helpful (i.e., not a, b, c, but real descriptors of what the thing is)
  
## 4. Calling functions
  + R has several built-in functions
  + Sometimes a function you need will be in a specific package
      + Installing and Loading packages
      
```{r}

# function for only installing packages you don't have and loading libraries you need
source("custom_functions/install_needed_pkgs.R")

# character vector for needed packages
needed_packages <- c("tidyverse")

install_needed_pkgs(needed_packages)

# sets the functions as being part of the packages you actually want to pull from
filter <- dplyr::filter
tibble <- tibble::tibble

```

  + ?function will bring up a help screen for that function
      + Can see what the arguments are that you need
      + Can see the defaults for the arguments
  + Calling a function requires you type the name + the arguments it requires 
      + Demo: merge(df1, df2, by=”variable name”, all=TRUE)
      
```{r}

# create four variables: N, id, cond x, z, and y
# 10 observations per variable
N <- 10

# create a condition variable
cond <- rep(c("Treatment", "Control"), each = 5)

# let's just try out rep function this other way
rep(c("Treatment", "Control"), 2)

# set randomizer seed so results can be reproduced
set.seed(1234)

# drawn from random normal distribution
gratitude_t1 <- rnorm(n = N, mean = 4.5, sd = 0.95)
gratitude_t2 <- rnorm(n = N-1, mean = 5.5, sd = 0.95)
gratitude_t2 <- c(gratitude_t2, NA)
interdep <- 0.8 * gratitude_t1 + rnorm(n = N, mean = 0, sd = 0.5)

# store in a dataframe
time1_survey <- tibble(id = 1:10, cond, gratitude_t1, interdep)
time2_survey <- tibble(id = c(1:9, NA), gratitude_t2)


# coerce condition to be a factor
#I will talk about mutate again later

time1_survey <-
  time1_survey %>%
  mutate(cond = factor(cond))

```
    
    + Demo: join(x, y, by)

```{r}

# join time 1 and time 2 by subject id
# ?left_join (Join matching rows from y to x)
# right_join
# ?full_join

full_joined_survey <- full_join(x = time1_survey, y = time2_survey, by = "id")

joined_survey <- inner_join(x = time1_survey, y = time2_survey, by = "id")

left_joined_survey <- left_join(x = time1_survey, y = time2_survey, by = "id")

right_joined_survey <- right_join(x = time1_survey, y = time2_survey, by = "id")

# most of the time you want to use left_join, full_join, (sometimes inner_join)

```
      
      + Demo: scale(variable, scale=FALSE)
        + scale will center your variable if the argument scale is FALSE
        + it will z-score your variable if the argument scale is TRUE
        
```{r}

# ?scale
joined_survey <- joined_survey %>%
  mutate(interdep_c = parse_number(scale(interdep, center = TRUE, scale = FALSE)))

```

## 5. Useful functions for data management/transformation
  + filter()
    + Remember that to save these things, we need to assign it to a new object (usually I try not to update the original dataframe too much)
  
```{r}

# filter only id "exactly equal to" 9
# need to use == b/c we're using logical (Boolean) language
filter(joined_survey, id == 9)

# filter only id "less than" 7
filter(joined_survey, id < 7)

# filter only id "less than or equal to" 7
filter(joined_survey, id <= 7)

# filter only ids 1 or 2 or 7
filter(joined_survey, id %in% c(1, 2, 7))

# filter only id "less than" 7 "or" id "exactly equal to" 9
filter(joined_survey, id < 7 | id == 9)

# filter only id "less than" 9 "and" gratitude_t1 "greater than" 4
filter(joined_survey, id < 7 & gratitude_t1 > 4)

# filter only condition "not equal to" Treatment, in a "tidy way"
joined_survey %>%
  filter(cond != "Treatment")

#filter out NAs for specific variable (I chose id)
time2_survey

time2_survey %>%
  filter(!is.na(id))

```
  
  + arrange()
    + arrange allows you to sort the data based on certain columns
    
```{r}

# arrange joined_survey by cond
arrange(joined_survey, cond)

# arrange joined_survey by id
arrange(joined_survey, id)

# arrange joined_survey by cond, then interdependence
arrange(joined_survey, cond, interdep)

```
  
  + select()
    + select allows you to select specific columns (for instance if you're working with a giant dataframe)
    
```{r}

# from joined_survey, select id, interdep_c, and gratitude_t1
select(joined_survey, id, interdep_c, gratitude_t1)

#a more tidy way
joined_survey %>%
  select(id, interdep_c, gratitude_t1)

```
  
  + mutate()
    + mutate allows you to create new variables
    
```{r}

# mutate gratitude_mean
# two ways to do it
# note: not saving mutated variables; to do so, must save to old or new object (joined_survey) with <-
mutate(joined_survey,
       gratitude_mean = (gratitude_t1 + gratitude_t2) / 2,
       gratitude_mean_alt = mean(c(gratitude_t1, gratitude_t2),
                                 na.rm = TRUE))

# using pipes
# have to use rowwise
joined_survey %>% 
  rowwise() %>% 
  mutate(gratitude_mean = (gratitude_t1 + gratitude_t2) / 2, 
         gratitude_mean_alt = mean(c(gratitude_t1, gratitude_t2), na.rm = TRUE)) %>%
  ungroup()

```
  
  + summarise()
    + summarise() will also allow you to create variables at the total group level
    
```{r}

summarise(joined_survey, mean(x = c(gratitude_t1, gratitude_t2), na.rm = TRUE))

#group by

joined_survey %>%
  group_by(cond)%>%
  summarise(mean_grat = mean(gratitude_t1, na.rm = TRUE))

```
   
   + gather()
    + puts data into long format
    
```{r}

# variable name (key) = time and response value (value) = gratitude
# arrange by Subj ID
joined_survey_long <- gather(data = joined_survey, key = time, value = gratitude_score, gratitude_t1, gratitude_t2) %>%
  arrange(id)

```
 
 + spread() 
    + puts data into wide format
    
```{r}

joined_survey_wide <- spread(data = joined_survey_long, key = time, value = gratitude_score)

```
  
  + ifelse()
    + the same as the if function in excel, basically says, "if statement is true, return this value, otherwise return this other value."
    
```{r}

# in a tidyverse way
joined_survey <- 
  joined_survey %>%
  mutate(cond2 = ifelse((cond == "Treatment" & id <= 3) | (cond == "Control" & id <= 7), "Family",
                 ifelse((cond == "Treatment" & id <= 5) | (cond == "Control" & id >= 8), "Friends",  NA)))

```
