---
title: "exercise: academic salaries"
output: 
  html_document: 
    toc: true
    toc_float: true
    fig_height: 7.5
    fig_width: 10.5
---

# install packages and load their libraries
> Load the libraries you need for this session.  
> Include code that installs those packages for a hypothetical user who hasn't installed those packages (make sure to # comment out the code before knitting this document)

```{r, message = FALSE, warning = FALSE}

```

# read data
> read the data in the file called "C04e01dt1.txt" and save it to an object called "publications"
> there are no variable names, so use an argument that tells the function not to use the first row of data to make column/variable names

```{r, message = FALSE, warning = FALSE}

```

## add variable names
> 1. id1 (case ID #1)
> 2. id2 (case ID #2)  
> 3. yearsPhD (years since PhD)  
> 4. publications (publication count)  
> 5. man (dummy coded gender; 0 = woman, 1 = man)  
> 6. citations (citation count)  
> 7. salary (salary) 

```{r}

```

## transform id1 and id2 into factor variables
> as is, they are numeric variables

```{r}

```

## recode man so woman = -0.5 and man = 0.5

```{r}

```

## create new variable called gender_lbl
> labels: Woman, Man

```{r}

```

# plot a scatterplot matrix of years since PhD, publications, citations, and salary

```{r}

```

# plot boxplots of years since PhD, publications, citations, and salary by gender

```{r}

```

# test whether men and women have equal average salaries

```{r}

```

# estimate the relationship between publications and salary
> report the estimate in salary units  
> what does the intercept mean?

## fit the model

```{r}

```

## generate diagnostic plots
> Is there a relationship between predicted values and residuals? (i.e., is there a pattern in the first plot?)  
> Do the resduals look normal? (i.e., the dots "hug" the line in the second plot)  

```{r}

```

## print results

```{r}

```

# visualize the relationship between publications and salary by gender

```{r}

```

## mean center the publications variable
> subtract the mean of publications from each publications value

```{r}

```

## replot the relationship with the centered publication variable

```{r}

```

# estimate the relationship between publications and salary and whether it depends on gender
> use the mean-centered publications variable  
> report the estimate in salary units  
> what does the coefficient for gender mean?  
> what does the intercept mean?  

## fit the model

```{r}

```

## generate diagnostic plots
> Is there a relationship between predicted values and residuals? (i.e., is there a pattern in the first plot?)  
> Do the resduals look normal? (i.e., the dots "hug" the line in the second plot)  

```{r}

```

## save the cook's distance values to your dataset
> 1. create a new variable in dataset called "cooks.d2" (i.e., cook's distance values for the second model)  
> 2. see help("cooks.distance")

```{r}

```

## plot cook's distance values on the y-axis and id1 on the y-axis
> cook's distance indexes the influence of an observation on the regression: bigger values mean including that observation had a bigger effect on the regression results  
> Looking at the plot, are there any influential observations?  

```{r}

```

## print results

```{r}

```
