---
title: "Exercise Key: Academic Salaries"
date: "06/20/2019"
output: 
  html_document: 
    fig_height: 7.5
    fig_width: 10.5
    theme: flatly
    toc: yes
    toc_float: yes
---

# Install packages and load their libraries
> Load the libraries you need for this session.  
> Include code that installs those packages for a hypothetical user who hasn't installed those packages (make sure to # comment out the code before knitting this document)

```{r, message = FALSE, warning = FALSE}

# install.packages("tidyverse")
# install.packages("psych")

library(tidyverse)
library(psych)

```

# Read data
> Read the data in the file called "C04e01dt1.txt" and save it to an object called "publications"
> There are no variable names, so use an argument that tells the function not to use the first row of data to make column/variable names

```{r, message = FALSE, warning = FALSE}

publications <- read_table(file = "data/C04e01dt1.txt", col_names = FALSE)

```

## Add variable names
> 1. id1 (case ID #1)
> 2. id2 (case ID #2)  
> 3. yearsPhD (years since PhD)  
> 4. publications (publication count)  
> 5. man (dummy coded gender; 0 = woman, 1 = man)  
> 6. citations (citation count)  
> 7. salary (salary) 

```{r}

colnames(publications) <- c("id1", "id2", "yearsPhD", "publications", "man", "citations", "salary")

```

## Transform id1 and id2 into factor variables
> As is, they are numeric variables.

```{r}

publications$id1 <- factor(publications$id1)
publications$id2 <- factor(publications$id2)

```

## Recode man so woman = -0.5 and man = 0.5

```{r}

publications$man <- with(data = publications, recode(man, "0" = -0.5, "1" = 0.5))

```

## Create new variable called gender_lbl
> Labels: Woman, Man

```{r}

publications$gender_lbl <- with(data = publications, recode(man, "-0.5" = "Woman", "0.5" = "Man"))

```

# Plot a scatterplot matrix of years since PhD, publications, citations, and salary

```{r}

publications %>% 
  select(yearsPhD, publications, citations, salary) %>% 
  pairs.panels()

```

# Plot boxplots of years since PhD, publications, citations, and salary by gender

```{r}

publications %>% 
  gather(key = variable, value = value, yearsPhD, publications, citations, salary) %>% 
  ggplot(mapping = aes(x = gender_lbl, y = value)) +
  geom_boxplot() +
  facet_wrap(facets = ~ variable, scales = "free")

```

# Test whether men and women have equal average salaries

```{r}

t.test(salary ~ gender_lbl, data = publications)

```

# Estimate the relationship between publications and salary
> Report the estimate in salary units  
> What does the intercept mean?

## Fit the model

```{r}

lm.fit1 <- lm(salary ~ publications, data = publications)

```

## Generate diagnostic plots
> Is there a relationship between predicted values and residuals? (i.e., is there a pattern in the first plot?)  
> Do the resduals look normal? (i.e., the dots "hug" the line in the second plot)  

```{r}

plot(lm.fit1)

```

## Print results

```{r}

summary(lm.fit1)

```

# Visualize the relationship between publications and salary by gender

```{r}

publications %>% 
  ggplot(mapping = aes(x = publications, y = salary, color = gender_lbl)) +
  geom_point() +
  geom_smooth(method = "lm")

```

## Mean center the publications variable
> Subtract the mean of publications from each publications value.

```{r}

publications$publications_c <- with(data = publications, publications - mean(publications))

```

## Replot the relationship with the centered publication variable

```{r}

publications %>% 
  ggplot(mapping = aes(x = publications_c, y = salary, color = gender_lbl)) +
  geom_point() +
  geom_smooth(method = "lm")

```

# Estimate the relationship between publications and salary and whether it depends on gender
> Use the mean-centered publications variable  
> Report the estimate in salary units  
> What does the coefficient for gender mean?  
> What does the intercept mean?  

## Fit the model

```{r}

lm.fit2 <- lm(salary ~ publications_c * man, data = publications)

```

## Generate diagnostic plots
> Is there a relationship between predicted values and residuals? (i.e., is there a pattern in the first plot?)  
> Do the resduals look normal? (i.e., the dots "hug" the line in the second plot)  

```{r}

plot(lm.fit2)

```

## Save the cook's distance values to your dataset
> 1. Create a new variable in dataset called "cooks.d2" (i.e., cook's distance values for the second model)  
> 2. See help("cooks.distance")

```{r}

publications$cooks.d2 <- cooks.distance(lm.fit2)

```

## Plot cook's distance values on the y-axis and id1 on the y-axis
> Cook's distance indexes the influence of an observation on the regression: bigger values mean including that observation had a bigger effect on the regression results  
> Looking at the plot, are there any influential observations?  

```{r}

publications %>% 
  ggplot(mapping = aes(x = id1, y = cooks.d2)) +
  geom_point() +
  coord_flip()

```

## Print results

```{r}

summary(lm.fit2)

```
