---
title: "study 1 data analysis"
date: "06/06/2018"
---

# packages

```{r, warning = FALSE, message = FALSE}

# install.packages("tidyverse")
# install.packages("knitr")
# install.packages("psych")
# install.packages("car")
# install.packages("heplots")
library(tidyverse)
library(knitr)
library(psych)
library(car)
library(heplots)

# I want recode from dplyr
recode <- dplyr::recode

```

# data

```{r, warning = FALSE, message = FALSE}

hsb2 <- "data/study1/cleandata/hsb2.csv" %>% read_csv()

```

# custom functions

```{r}

source("functions/oneway.R")

```

# create contrast variables

```{r}

hsb2 <- hsb2 %>%
  mutate(lowhigh = recode(ses_lbl, "Low" = -0.5, "Middle" = 0, "High" = 0.5),
         lomidVhi = recode(ses_lbl, "Low" = -0.25, "Middle" = -0.25, "High" = 0.5),
         acadgenrl = recode(program_lbl, "Academic" = -0.5, "General" = 0.5, "Vocation" = 0),
         vocVacgen = recode(program_lbl, "Academic" = -0.25, "General" = -0.25, "Vocation" = 0.5),
         ses_program = interaction(ses_lbl, program_lbl, sep = " x ") %>% factor(levels = ))

hsb2 %>% 
  count(ses_program, ses_lbl, program_lbl, lowhigh, lomidVhi, acadgenrl, vocVacgen) %>% 
  kable()

```

# model

## does social studies score depend on socioeconomic status and program?

### fit linear regression

```{r}

lm1 <- lm(SOCST ~ (lowhigh + lomidVhi) * (acadgenrl + vocVacgen), data = hsb2)

```

### results (specific contrasts)

```{r}

summary(lm1)

```

### ANOVA results

```{r}

Anova(lm1, type = "3")

```

### eta-squared and partial eta-squared

```{r}

etasq(lm1, partial = FALSE)
etasq(lm1, partial = TRUE)

```

## Welch contrasts
> see [Testing Between-Subjects Contrasts in R
](http://nickmichalak.com/blog_entries/2018/nrg06/nrg06.html)

```{r}

with(hsb2, oneway(dv = SOCST, group = ses_program, contrast = list(lowhigh = c(1, -1, 0, 1, -1, 0, -1, 1, 0)), alpha = 0.05))

```

# Does reading aptidude supress, mediate, or confound the relationship between math and science aptitude?
> MacKinnon, D. P., Krull, J. L., & Lockwood, C. M. (2000). Equivalence of the mediation, confounding and suppression effect. Prevention science, 1(4), 173-181. [.pdf](http://w3dev.ualg.pt/~lfaisca/SMAD05/Prevention%20Science_Mackinnon_2000.pdf)

## save math only and math + science models

```{r}

lm2 <- lm(SCIENCE ~ MATH, data = hsb2)
lm3 <- lm(SCIENCE ~ MATH + READ, data = hsb2)

```

### examine diagnostic plots for the full model

```{r}

plot(lm3)

```

### results for both models

```{r}

summary(lm2)
summary(lm3)

```

### (adjusted) R-squared change

```{r}

summary(lm3)$adj.r.squared - summary(lm2)$adj.r.squared

```

### test (adjusted) R-squared change

```{r}

anova(lm2, lm3)

```

### simple indirect effect test

```{r}

mediate(SCIENCE ~ MATH + (READ), data = hsb2, n.iter = 10000, use = "pairwise", plot = TRUE) %>% 
  print(short = FALSE)

```





